# Linear models {#sec:linmod}

The objective of this chapter is to provide students with an intuitive
understanding of how linear models are used in significance testing,
and more specifically

- show how linear models are equivalent to a t-test
- demonstrate their flexibility.

## Starting with a t-test

`r msmbstyle::question_begin()`

Load the `gexp` data from the `rWSBIM2122` package (version >= 0.2.2)
as show below. It provides the log2 expression of a single gene of
interest measured in two groups of interest.

```{r gexp}
library(rWSBIM2122)
data(gexp)
gexp
```

Using a t-test, compare the expression in groups 1 and 2. In
particular, report the mean expression of each group, the log2
fold-change, the associated p-value and visualise the data.

`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`

```{r gexp_t_test1}
t.test(expression ~ group, data = gexp)
```

```{r gexp_vis1, message = FALSE}
library("ggplot2")
p <- ggplot(gexp, aes(x = group, y = expression)) +
    geom_boxplot() + geom_jitter()
p
```

`r msmbstyle::solution_end()`

## The linear model equivalence


Let's now use linear regression and the `lm()` function to model the
expression in groups 1 and 2:

$$y = \beta_0 + \beta_1 \times x + \epsilon$$

where $y$ is the expression we want to mode and $x$ represents the two
groups that are generally encode as 0 and 1. The expression above
translates as follows:

- the expression in group 1 (i.e. when $x = 0$) is equal to $\beta_0$
  (the average expression of group 1);
- the expression in group 2 (i.e. when $x = 1$) is equal to
  $\beta_0$ + $\beta_1$ (the difference between group 1 and group 2).

We model this with

```{r mod1}
mod1 <- lm(expression ~ group, data = gexp)
mod1
```

An indeed, we see that

1. the intercept corresponds to the mean in group 1, and 
2. the slope (`group2` coefficient) of the model corresponds to the
   log2 fold-change and that, as expected, the mean of group 2 is
   *intercept + slope*.

We can visualise this linear model with:

```{r}
p  +
    geom_point(data = data.frame(x = c(1, 2), y = c(2.75, 4.33)),
               aes(x = x, y = y),
               colour = "red", size = 5) +
    geom_abline(intercept = coefficients(mod1)[1] - coefficients(mod1)[2],
                slope = coefficients(mod1)[2])
```

Note that above, we need to subtract the slope from the intercept when
plotting the linear model because groups 1 and 2 are encoded as 0 and
1 in the model, but plotted as 1 and 2 on the figure.

We can also extract the p-value associated with the coefficients:

```{r}
summary(mod1)
```

that are computer as 

$$t_{score} = \frac{\beta_i - 0}{SE_{\beta_i}}$$

that has a t-distribution with n − 2 degrees of freedom if the null
hypothesis is true. We use a t-test that tests whether the slope is
different from 0: 

- $H_0$: $\beta_i = 0$ 
- $H_1$: $\beta_i \neq 0$ 

The p-values for the slope $\beta_1$ is slightly different due to the
heteroscedasticity assumption that wasn't set in the t-test, which we
can repeat:


```{r}
t.test(expression ~ group, data = gexp, var.equal = TRUE)
```


## Paired t-test

`r msmbstyle::question_begin()`

Repeat the test comparing the gene expression between group 1 and
group 2 using a *paired* t-test. Report the mean expression of each
group, the log2 fold-change, the associated p-value and visualise the
data.

`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`

```{r gexp_t_test2}
t.test(expression ~ group, data = gexp, paired = TRUE)
```

```{r gexp_vis2}
gexp2 <- dplyr::full_join(gexp[1:10, ], gexp[11:20, ],
                          by = "ID", suffix = c("_1", "_2"))

ggplot(gexp, aes(x = group, y = expression)) +
    geom_boxplot() + 
    geom_point(aes(colour = ID), size = 4) +
    geom_segment(data = gexp2,
                 aes(x = group_1, xend = group_2,
                     y = expression_1, yend = expression_2,
                     colour = factor(1:10)))
```


`r msmbstyle::solution_end()`

This comparison can be replicated using a linar model 

$$y = \beta_0 + \beta_1 \times group + \beta_2 \times ID + \epsilon$$

where the $\beta_2$ coefficient will encode the relation between pairs
of individuals. We model this with

```{r mod2}
mod2 <- lm(expression ~ group + ID, data = gexp)
mod2
```

The `group2` coefficient still corresponds to the fold-change between
the two groups. The coefficients `ID2` to `ID10` correspond to the
respective differences between the average expression of `ID1` and
`ID2` to `ID10`. 

```{r}
avg_id <- (gexp2$expression_1 + gexp2$expression_2)/2
coefs_id <- avg_id[-1] - avg_id[1]
names(coefs_id) <- paste0("ID", 2:10)
coefs_id
```

As above, a summary of the model will give us the statistical
significance for the different coefficients.

```{r}
summary(mod2)
```

`r msmbstyle::question_begin()`

Interpret the summary of `mod2` computed above.

`r msmbstyle::question_end()`

`r msmbstyle::solution_begin()`

We see that the `group2` coefficient, that tests the difference
between the two groups is now significant. It is identical to the
paired t-test above.


The respective ID coefficients compare the average expression in these
IDs agains the average expression of ID1. We can indeed see that ID6
and ID7, that have significant coefficients, are the most different
from ID1.

`r msmbstyle::solution_end()`

## Model generalisation

In the example above, we had two variables that influenced the gene
expression, namely the group and the specific individual and we ended
up with a model that included both of these[^tilda]:

[^tilda]: We will drop the left-hand side term of our formula, setting
    it by default as the quantiative expression data.


```{r, eval = FALSE}
~ group + ID
```

Any additional variables could be added to the model if deemed necessary.


`r msmbstyle::question_begin()`

We have measured the gene expression in cell lines U2OS and Jurkat
(defined as variable `cell_line`), each in the presence of drug A or B
(defined as variable `drug`). How would you define the test formula?

`r msmbstyle::question_end()`

`r msmbstyle::question_begin()`

As above, we have measured the gene expression in cell lines U2OS and
Jurkat (defined as variable `cell_line`), each in the presence of drug
A or B (defined as variable `drug`). This time however, two operators
(defined as variable `operator`) have shared the work. How would you
define the test formula?

`r msmbstyle::question_end()`


`r msmbstyle::question_begin()`

We have measured the gene expression in cell lines U2OS and Jurkat
(defined as variable `cell_line`), each in the presence of drug A, B
or C (defined as variable `drug`). How would you define the test
formula? How many coefficients will you have? Describe them.

`r msmbstyle::question_end()`


## Interactions between variables

