# More complex experimental design {#sec-design}

**Learning Objectives**

The goal of this chapter is

- to understand more complex experimental designs
- how to extract relevant factors

```{r, echo = FALSE}
suppressPackageStartupMessages(library("DESeq2"))
suppressPackageStartupMessages(library("tidyverse"))
```

## A more complex dataset


The dataset that we used previously was quite simple, we had 3 KD
samples vs 3 control samples.  But the original dataset was in reality
a bit more complex, as the KD experiments were actually done on 2
different cell lines.  So we have now 12 samples. The 6 first samples
correspond to an epithelial cell line (Cell1), and 6 other samples
correspond to a fibroblast cell line (Cell2).

```{r, echo=FALSE}
stopifnot(packageVersion("rWSBIM2122") >= "0.4.1")
```

```{r}
load("wsbim2122_data/deseq2/coldata_both.rda")
load("wsbim2122_data/deseq2/counts_both.rda")
coldata
```

### PCA analysis

Let's start by a PCA analysis, that will help us to define the most
appropriate design to use for the analysis.

To generate the [DESeqDataSet](https://www.rdocumentation.org/packages/DESeq2/versions/1.12.3/topics/DESeqDataSet-class), remember that we have to specify the experimental design.

We will see that can use different experimental designs for the analysis,
depending on the biological questions that we want to answer. We will discuss
about this latter. At this step, as we just want to do a PCA (that will anyway
be blind to the sample information specified by the design formula),
we can use a design ~ 1.

```{r, message = FALSE, warning = FALSE}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design =  ~ 1)
dds <- DESeq(dds)

rld <- rlogTransformation(dds)
pca_data <- plotPCA(rld,
                    intgroup = c("Condition", "Type"),
                    returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))
ggplot(pca_data,
       aes(x = PC1, y = PC2,
           color = Condition, shape = Type)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance"))
```

We can see from this PCA plot that PC1 separates the samples based on the
cell line origin, while PC2 separates the samples based on the siRNA treatment.
This means that most of the variability comes from the cell origin.

The design used for the analysis will have to take this cell effect into
account.

Different designs can be used to analyse this dataset, depending on the
biological questions that are asked.

- **A paired design** should be used to answer to the question:

*Which genes are consistently affected by the KD in both cell lines?*

- **A design with an interaction** should be used if the biological questions are:

*Which genes are significantly affected by the KD in the epithelial cells?*

*Which genes are significantly affected by the KD in the fibroblasts?*

*Which genes are not affected in the same way by the KD in the two cell lines*


## Paired design

To take into account the condition effect while controlling the cell origin,
we can use the following design:

`design = ~ Cell + Condition`

Let's run the analysis with this design (we have to re-run the
`DESeqDataSetFromMatrix()` and `DESeq()` functions)

```{r, message = FALSE, warning = FALSE}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design =  ~ Type + Condition)

# Set the mock level as the reference level
dds$Condition <- relevel(dds$Condition, ref = "mock")
# Set the Epithelial cells as the reference level
dds$Type <- relevel(dds$Type, ref = "Epithelial")

dds <- DESeq(dds)
```

As we have seen previously, the `resultsNames()` function gives the
names of the coefficients that can be extracted.

```{r}
resultsNames(dds)
```


The linear model set with this design can be written

$$log2(q_{ij}) = Intercept + Type\_Fibroblast\_vs\_Epithelial.x_j + Condition\_KD\_vs\_mock.y_j + \epsilon$$

- The $Intercept$ represents the log2 expression level in the
reference (in mock Epithelial cells).

- The $Type\_Fibroblast\_vs\_Epithelial$ coefficient corresponds to the log2FC
between fibroblasts and epithelial cells. $x_j = 0$ if the sample j corresponds to epithelial cells,
and $x_j = 1$ if the sample j corresponds to fibroblasts.

- The $Condition\_KD\_vs\_mock$ coefficient corresponds to the log2FC between KD and
mock cells. $y_j = 0$ if the sample j corresponds
to mock cells, and $y_j$ = 1 if the sample j corresponds to KD cells.


The following figure (generated with the [ExploreModelMatrix](http://www.bioconductor.org/packages/release/bioc/vignettes/ExploreModelMatrix/inst/doc/ExploreModelMatrix.html) package) illustrates the value of the
linear predictor of a generalized linear model for each combination of
input variables. It can help to understand and to generate contrasts.

```{r echo=FALSE}
coldata$Condition <- factor(coldata$Condition, levels = c("mock", "KD"))
coldata$Type <- factor(coldata$Type, levels = c("Epithelial", "Fibroblast"))
vd <- ExploreModelMatrix::VisualizeDesign(sampleData = coldata,
                      designFormula = ~ Type + Condition,
                      textSizeFitted = 6)
vd$plotlist[[1]]
```

From this figure, we can see how the genes log2 expression values are modelized
in the different samples.

- In mock epithelial cells
$log2(q_{Gene_i\_Epithelial\_mock}) = Intercept$

- In KD epithelial cells
$log2(q_{Gene_i\_Epithelial\_KD}) = Intercept + Condition\_KD\_vs\_mock$

- In mock fibroblasts cells
$log2(q_{Gene_i\_Fibroblast\_mock}) = Intercept + Type\_Fibroblast\_vs\_Epithelial$

- In KD fibroblasts cells
$log2(q_{Gene_i\_Fibroblast\_KD}) = Intercept + Type\_Fibroblast\_vs\_Epithelial + Condition\_KD\_vs\_mock$

### KD effect

To identify the genes that are consistently affected by the KD in both cell lines,
we will have to test (for each gene) the *Condition_KD_vs_mock* coefficient, to see
if it is significantly different from zero.

Remember that we can use the `name` parameter of the `results()` function to specify the
coefficient that we want to extract (here the *Condition_KD_vs_mock* coefficient).
This will test if genes are differentially expressed in KD cells versus
mock cells.

```{r}
res <- results(dds, name = "Condition_KD_vs_mock")
res_tbl <- as_tibble(res, rownames = "ENSEMBL") %>%
  arrange(padj)
head(res_tbl)
```

The normalised counts of the genes with the lowest p-adjusted values are plotted bellow.
We can see that these genes correspond to genes that are highly affected by the siRNA
in both cells lines.

```{r}
as_tibble(counts(dds[res_tbl$ENSEMBL[1:2], ], normalize = TRUE),
          rownames = 'ENSEMBL') %>%
  pivot_longer(names_to = "sample", values_to = "counts", -ENSEMBL) %>%
  left_join(as_tibble(colData(dds), rownames = "sample")) %>%
  mutate(name = paste0(substr(Type, 1, 5), '_', Condition, '_', 1:3)) %>%
  ggplot(aes(x = name, y = counts, fill = Condition)) +
  geom_bar(stat = 'identity', color = "gray30") +
  facet_wrap( ~ ENSEMBL, scales = "free") +
  theme(axis.text.x = element_text(size = 8, angle = 90),
        axis.title.x = element_blank(),
        legend.position = "right",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7))
```


### Epithelial vs fibroblasts cells

Of course if the question of interest was

*What are the genes that are differentially expressed in epithelial and fibroblasts?*

we would have tested the *Type_Fibroblast_vs_Epithelial* coefficient.


```{r}
res <- results(dds, name = "Type_Fibroblast_vs_Epithelial")
res_tbl <- as_tibble(res, rownames = "ENSEMBL") %>%
  arrange(padj)
as_tibble(counts(dds[res_tbl$ENSEMBL[1:2], ], normalize = TRUE),
          rownames = 'ENSEMBL') %>%
  pivot_longer(names_to = "sample", values_to = "counts", -ENSEMBL) %>%
  left_join(as_tibble(colData(dds), rownames = "sample")) %>%
  mutate(name = paste0(substr(Type, 1, 5), '_', Condition, '_', 1:3)) %>%
  ggplot(aes(x = name, y = counts, fill = Condition)) +
  geom_bar(stat = 'identity', color = "gray30") +
  facet_wrap( ~ ENSEMBL, scales = "free") +
  theme(axis.text.x = element_text(size = 8, angle = 90),
        axis.title.x = element_blank(),
        legend.position = "right",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7))
```






## Design with interaction

Interaction terms can be added to the design formula, in order to test
the treatment effect in one cell line or the treatment
effect that differs across the cell lines.

The design with an interaction could be written $$design = ~ Cell * Condition$$
which is equivalent to
$$design = ~ Cell + Condition + Cell:Condition$$


Let's use this design in our analysis (we have to re-run the
`DESeqDataSetFromMatrix()` and `DESeq()` functions)

```{r, message = FALSE, warning = FALSE}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design =  ~ Type * Condition)
# Set the Epithelial cells as the reference level
dds$Type <- relevel(dds$Type, ref = "Epithelial")
# Set the mock condition as the reference level
dds$Condition <- relevel(dds$Condition, ref = "mock")
dds <- DESeq(dds)
```

The `resultsNames()` function gives the names of results that can be extracted.

```{r}
resultsNames(dds)
```


The linear model used with this design can be written

$$log2(q_{ij}) = Intercept + Type\_Fibroblast\_vs\_Epithelial.x_j + Condition\_KD\_vs\_mock.y_j  + TypeFibroblast.ConditionKD.z_j + \epsilon$$

- The $Intercept$ represents the log2 expression level in the
reference (in mock Cell1).

- The $Type\_Fibroblast\_vs\_Epithelial$ coefficient corresponds
to the log2FC between mock fibroblasts and mock epithelial cells.
$x_j = 0$ if the sample j corresponds to epithelial cells,
and $x_j = 1$ if the sample j corresponds to fibroblasts.

- The $Condition\_KD\_vs\_mock$ coefficient corresponds to the log2FC
between KD and mock cells in the reference cells (here epithelial cells).
$y_j = 0$ if the sample j corresponds to mock cells, and $y_j$ = 1 if
the sample j corresponds to KD cells.

- The $TypeFibroblast.ConditionKD$ coefficient corresponds to the
the *extra KD effect* in KD-fibroblasts compared to KD-epitelial cells.
$z_j = 0$ if the sample j corresponds to mock cells or to epithial cells,
and $z_j = 1$ if the sample j corresponds to KD-fibroblasts.

This is summarized in the following figure:

```{r echo=FALSE}
coldata$Condition <- factor(coldata$Condition, levels = c("mock", "KD"))
coldata$Type <- factor(coldata$Type, levels = c("Epithelial", "Fibroblast"))
vd <- ExploreModelMatrix::VisualizeDesign(sampleData = coldata,
                      designFormula = ~ Type * Condition,
                      textSizeFitted = 4)
vd$plotlist[[1]]
```

From this figure, we can see how the genes log2 expression values
are modelized in the different samples.

- in mock epithelial cells
$log2(q_{Gene_i\_Epithelial\_mock}) = Intercept$

- in KD epithelial cells
$log2(q_{Gene_i\_Epithelial\_KD}) = Intercept + Condition\_KD\_vs\_mock$

- in mock fibroblasts cells
$log2(q_{Gene_i\_Fibroblast\_mock}) = Intercept + Type\_Fibroblast\_vs\_Epithelial$

- in KD fibroblasts cells
$log2(q_{Gene_i\_Fibroblast\_KD}) = Intercept + Type\_Fibroblast\_vs\_Epithelial + Condition\_KD\_vs\_mock + TypeFibroblast.ConditionKD$


### KD effect in Epithelial cells

As the epithelial cells were set as the reference cells,
the *Condition_KD_vs_mock* coefficient should be extracted to
test the effect of the KD in these cells.

In fact, testing the siRNA effect in epithelial cells is like comparing the
logFC between epithelial KD and epithelial mock cells.

$$log2(\frac{q_{Gene_i\_Epithelial\_KD}}{q_{Gene_i\_Epithelial\_mock}})$$

$$= log2(q_{Gene_i\_Epithelial\_KD}) - log2(q_{Gene_i\_Epithelial\_mock})$$
$$= Intercept + Condition\_KD - Intercept = Condition\_KD$$

```{r}
res <- results(dds, name = "Condition_KD_vs_mock")
res_tbl <- as_tibble(res, rownames = "ENSEMBL") %>%
  arrange(padj)
head(res_tbl)

as_tibble(counts(dds[res_tbl$ENSEMBL[1:2], ], normalize = TRUE),
          rownames = 'ENSEMBL') %>%
  pivot_longer(names_to = "sample", values_to = "counts", -ENSEMBL) %>%
  left_join(as_tibble(colData(dds), rownames = "sample")) %>%
  mutate(name = paste0(substr(Type, 1, 5), '_', Condition, '_', 1:3)) %>%
  ggplot(aes(x = name, y = counts, fill = Condition)) +
  geom_bar(stat = 'identity', color = "gray30") +
  facet_wrap( ~ ENSEMBL, scales = "free") +
  theme(axis.text.x = element_text(size = 8, angle = 90),
        axis.title.x = element_blank(),
        legend.position = "right",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7))
```

### KD effect in Fibroblasts cells

As the fibroblasts cells were NOT set as the reference cells,
it is less trivial to extract the right coefficients to test the
KD effect in fibroblasts.

But remember that testing the siRNA effect in fibroblast cells is like
comparing the logFC between fibroblast KD and fibroblast mock cells.

$$log2(\frac{q_{Gene_i\_Fibroblast\_KD}}{q_{Gene_i\_Fibroblast\_mock}})$$

$$= log2(q_{Gene_i\_Fibroblast\_KD}) - log2(q_{Gene_i\_Fibroblast\_mock})$$
$$= (Intercept + Condition\_KD + Type\_Fibroblast\_vs\_Epithelial +  TypeFibroblast.ConditionKD )$$ $$ - (Intercept + Type\_Fibroblast\_vs\_Epithelial) $$
$$= Condition\_KD +  TypeFibroblast.ConditionKD $$

So to test the KD effect in fibroblasts, we have to extract both
*Condition_KD_vs_mock* and *TypeFibroblast.ConditionKD* coefficients.

```{r}
res <- results(dds,
               list(c("Condition_KD_vs_mock", "TypeFibroblast.ConditionKD")))
res_tbl <- as_tibble(res, rownames = "ENSEMBL") %>%
  arrange(padj)
head(res_tbl)

as_tibble(counts(dds[res_tbl$ENSEMBL[1:2], ], normalize = TRUE),
          rownames = 'ENSEMBL') %>%
  pivot_longer(names_to = "sample", values_to = "counts", -ENSEMBL) %>%
  left_join(as_tibble(colData(dds), rownames = "sample")) %>%
  mutate(name = paste0(substr(Type, 1, 5), '_', Condition, '_', 1:3)) %>%
  ggplot(aes(x = name, y = counts, fill = Condition)) +
  geom_bar(stat = 'identity', color = "gray30") +
  facet_wrap( ~ ENSEMBL, scales = "free") +
  theme(axis.text.x = element_text(size = 8, angle = 90),
        axis.title.x = element_blank(),
        legend.position = "right",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7))
```


Note that another possibility would have been to set the fibroblasts
as the reference cells, re-run the whole analysis, and to simply
extract the *Condition_KD_vs_mock* coefficient.

```{r, message = FALSE, warning = FALSE}
dds_test <- DESeqDataSetFromMatrix(countData = counts,
                                   colData = coldata,
                                   design =  ~ Type * Condition)
# Set the Epithelial cells as the reference level
dds_test$Type <- relevel(dds_test$Type, ref = "Fibroblast")
# Set the mock condition as the reference level
dds_test$Condition <- relevel(dds_test$Condition, ref = "mock")
dds_test <- DESeq(dds_test)
res <- results(dds_test,
               name = "Condition_KD_vs_mock")
res_tbl <- as_tibble(res, rownames = "ENSEMBL") %>%
  arrange(padj)
head(res_tbl)
```

### KD effect that are different across epithelial and fibroblasts cells


Which genes are not affected in the same way by the KD in the two cell lines?
In other terms, we are testing the KD effects that are significantly different in both cell
lines. This can be done by extracting the *TypeFibroblast.ConditionKD* coefficient.
In fact this coefficient corresponds to the *extra KD effect* ocuring in fibroblasts
compared to epitelial cells.

If the effect of the KD is similar in both cell types, we expect the
*TypeFibroblast.ConditionKD* coefficient to be nearly 0.

If the effect of the KD is higher in fibroblasts than in epithelial cells, we expect the
*TypeFibroblast.ConditionKD* coefficient to be higher than 0.

If the effect of the KD is lower in fibroblasts than in epithelial cells, we expect the
*TypeFibroblast.ConditionKD* coefficient to be lower than 0.

```{r}
res <- results(dds, name = "TypeFibroblast.ConditionKD")
res_tbl <- as_tibble(res, rownames = "ENSEMBL") %>%
  arrange(padj)
head(res_tbl)
```

We can see that gene ENSG00000177494 (with the lowest p-value) has as
highly negative log2FC. This means that the effect of the siRNA treatment
is much lower in fibroblasts than in epithelial cells.

Gene ENSG00000108179 is also an illustrative example,
as its value is decreased by the siRNA treatment in the epithelial cells,
while on the contrary it is increased by the treatment in the fibroblast cells.


```{r}
as_tibble(counts(dds[res_tbl$ENSEMBL[c(1,5)], ], normalize = TRUE),
          rownames = 'ENSEMBL') %>%
  pivot_longer(names_to = "sample", values_to = "counts", -ENSEMBL) %>%
  left_join(as_tibble(colData(dds), rownames = "sample")) %>%
  mutate(name = paste0(substr(Type, 1, 5), '_', Condition, '_', 1:3)) %>%
  ggplot(aes(x = name, y = counts, fill = Condition)) +
  geom_bar(stat = 'identity', color = "gray30") +
  facet_wrap( ~ ENSEMBL, scales = "free") +
  theme(axis.text.x = element_text(size = 8, angle = 90),
        axis.title.x = element_blank(),
        legend.position = "right",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7))
```
